---
title: "EpiconceptIPD"
author: "Investigation team"
date: "04 Oktober 2018"
output: html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  echo = TRUE,
  oncordance = TRUE,
  message = FALSE,
  warning = FALSE
  )
```

```{r packages4knit, echo = FALSE}
#load packages
  #Those required for creating this markdown: 
    #"Worded": for styling and pagebreaks (devtools::install_github("davidgohel/worded"))
    #"knitr": for styling tables
required_packages <- c("worded", "knitr") 

for (i in seq(along = required_packages)) {
  library(required_packages[i], character.only = TRUE)
}

# The dataset for time series will probably come mid-next week. It will be Scottish data with several years of pre-vaccination data (2000-2006), 3 years of PCV7 years (2007-2009) and 8 years of PCV13 years (2010-2017). 
```

### Setting your working directory

You can check the path for your current working directory using the `getwd` function.

```{r, eval = FALSE}
# Check your current working directory
getwd()
```

To set your working directory you can use the `setwd` function. 

```{r, eval = FALSE}
setwd("C:/Users/Username/Desktop/EpiconceptIPD")
```

### Reading in files

Import the dataset from a comma seperated file (.csv) using the `read.csv` function, storing it as a dataframe within *R* called *ipd_cases*. Further, import population data using the same approach (*population*). For a CSV file the separator is normally a comma (","), however depending on the language of your operating system this can also be other values, for example some non-english speaking European countries use a semi-colon (";") as standard. Here we also specify that we do not want to format strings as factors. When reading population data we additionally define the option `check.names = FALSE` which makes sure that variable names are kept as defined in the CSV. By default (`check.names = TRUE`) *R* would automatically change column names to follow certain rules (e.g. no numeric as first character).

```{r read_data}
ipd_cases <- read.csv("data/ipd_cases.csv", sep = ",", stringsAsFactors = FALSE)
population <- read.csv("data/ipd_population.csv", sep = ",", stringsAsFactors = FALSE, check.names = FALSE)
```

Take a look at the first 10 rows of the data or browse the dataframe in the preview.

```{r, eval = FALSE}
head(ipd_cases, n = 10)
# or
View(ipd_cases)
```
Remember that `View` needs to be written with a capital *V*.

```{r, echo = FALSE}
kable(head(ipd_cases, n = 10))
```

Let us now define two variables which represent the serotypes covered by PCV7 and PCV13. We will use those variables later to identify cases with either so-called PCV7 or PCV13 serotypes respectively. We will e.g. call a case a "PCV7 case" if the identified serotype is one of the PCV7 serotypes.
```{r vaccine_definition}
pcv7  <- c("4", "6B", "9V", "14", "18C", "19F", "23F")
pcv13 <- c("4", "6B", "9V", "14", "18C", "19F", "23F", "1", "5", "7F", "3", "6A", "19A")
```


### Recoding/Fixing Data

```{r}
# decide if this should be taken out and a "clean" dataset should be available
ipd_cases$Month[!ipd_cases$Month %in% 1:12] <- NA

ipd_cases$Serotype <- gsub(pattern = "serotype ", replacement = "", x = ipd_cases$Serotype, fixed = TRUE)

```

### Basic statistics/analyses

Tabulate cases by year and age-group.
```{r}
all_year <- addmargins(table(ipd_cases$Year, ipd_cases$AgeGroup), 2, FUN = list(Total = sum)) # find better name for df
```

```{r, eval = FALSE}
View(all_year)
```

```{r, echo = FALSE}
kable(all_year)
```

Define two columns in our data to indicate if the case corresponds to PCV7 and/or PCV13 serotypes, respectively. Values in these columns are either `TRUE` or `FALSE`.
```{r}
ipd_cases$PCV7 <- ipd_cases$Serotype %in% pcv7
ipd_cases$PCV13 <- ipd_cases$Serotype %in% pcv13
```

Create tables of PCV7 (*pcv7_year*) and PCV13 (*pcv13_year*) cases by year and age-group. Remember, PCV7 was introduced in 2007, and PCV13 in 2010.
```{r}
pcv7_year <- addmargins(table(ipd_cases$Year[ipd_cases$PCV7], 
                              ipd_cases$AgeGroup[ipd_cases$PCV7]), 
                        margin = 2, FUN = list(Total = sum))
pcv13_year <- addmargins(table(ipd_cases$Year[ipd_cases$PCV13], 
                               ipd_cases$AgeGroup[ipd_cases$PCV13]), 
                         margin = 2, FUN = list(Total = sum))
```


Plot annual cases and highlight years with PCV7/13 introduction (red dashed vertical lines).

**PCV7:**
```{r}
plot(rownames(pcv7_year), pcv7_year[, "Total"], type = "l",
     xlab = "Year", ylab = "Number of cases",
     main = "IPD cases of PCV7 serotypes, total population",
     sub = "red dashed: PCV7 and PCV13 introduction, respectively",
     ylim = c(0, 450))
points(rownames(pcv7_year), pcv7_year[, "Total"])
abline(v = c(2007, 2010), col = "red", lty = "dashed")
```

**PCV13:**
```{r}
plot(rownames(pcv13_year), pcv13_year[, "Total"], type = "l",
     xlab = "Year", ylab = "Number of cases",
     main = "IPD cases of PCV13 serotypes, total population",
     sub = "red dashed: PCV7 and PCV13 introduction, respectively",
     ylim = c(0, 700))
points(rownames(pcv13_year), pcv13_year[, "Total"])
abline(v = c(2007, 2010), col = "red", lty = "dashed")
```

### Segmented regression
We will need the `glm.nb` function from the *MASS* package to perform negative binomial regression. Additionally we will use the `tidy` function from the *broom* package to create nice model outputs. If you did not yet install these packages, do so using the `install.packages` function.
```{r, eval = FALSE}
install.packages("MASS")
install.packages("broom")
```

Now load the required packages.
```{r}
library(MASS)
library(broom)
```


*blah blah, some basic information about segmented regression of ITS*

*Explain why we look at PCV13 IPD only and use PCV7 introduction as intervention (good to practice, easy enough, still big samplesize, 6 additional STs seem to not have that big of an impact, ...).*

We will analyse annual incidence of PCV13 serotypes among the whole population (*total*).

First we need to transform the table (*pcv13_year*) into a data.frame. We will do so by using the *R* function `as.data.frame.matrix`
Check the class of *pcv13_year*:
```{r}
class(pcv13_year)
```

In our case, the year is stored as row name and not as a column itself. Therefore transform the data into a data.frame as explained above and add the year as seperate column. Make sure that the year column is of class numeric.
```{r}
pcv13_aggr <- as.data.frame.matrix(pcv13_year)
pcv13_aggr$year <- as.numeric(rownames(pcv13_aggr))
```

Merge population and keep necessary columns only.
```{r}
pcv13_df <- merge(pcv13_aggr, population, by = "year", all.x = TRUE, suffixes = c("", "_pop"))
pcv13_df <- pcv13_df[, c("year", "Total", "Total_pop")]
```

Define parameters and columns needed for segmented regression (to be explained explicitely)

```{r}
intervention_year <- 2007

pcv13_df$linear <- 1:nrow(pcv13_df)
pcv13_df$const_vacc <- as.numeric(pcv13_df$year >= intervention_year)
pcv13_df$lin_vacc <- pmax(0, pcv13_df$year - intervention_year + 1)

pcv13_df$pop100 <- pcv13_df$Total_pop / 100000
```

Negative binomial segmented regression
```{r}
# negative binomial segmented regression
nb <- glm.nb(Total ~ linear + const_vacc + lin_vacc + offset(log(pop100)), data = pcv13_df)
```


Create table with results. Use the `tidy` function from *broom* package to simplify the regression output.
```{r}
nb_tidy <- tidy(nb, exponentiate = FALSE, conf.int = TRUE)
```

Take a look using
```{r, eval = FALSE}
View(nb_tidy)
```

```{r, echo = FALSE}
kable(nb_tidy)
```

Figure:

Prepare the data.frame for plotting. Calculate model predictions of incidence and confidence intervals.
```{r}
nb_predict <- pcv13_df
nb_predict$pop100 <- 1
pred.ci <- predict(nb, nb_predict, type = "link", se.fit = TRUE)
pred.ci <- within(pred.ci, {
  inci <- exp(fit)
  LL <- exp(fit - 1.96 * se.fit)
  UL <- exp(fit + 1.96 * se.fit)
})

nb_predict$inci_observed <- nb_predict$Total / nb_predict$Total_pop * 100000
nb_predict$inci_modelled <- pred.ci$inci
nb_predict$lowerCI <- pred.ci$LL
nb_predict$upperCI <- pred.ci$UL

# split into 2 data.frames for plotting. One for the pre- and the other for the post-period
pre <- nb_predict[nb_predict$year < intervention_year, ]
post <- nb_predict[nb_predict$year >= intervention_year, ]

```

```{r}
plot(nb_predict$year, nb_predict$inci_observed, type = "p",
     xlab = "Year", ylab = "Number of cases per 100.000 pop",
     main = "IPD incidence of PCV13 serotypes, total population",
     sub = "red dashed: PCV7 introduction",
     xlim = range(nb_predict$year),
     ylim = c(0, 15))

lines(pre$year, pre$inci_modelled, type = "l")
lines(pre$year, pre$lowerCI, type = "l", lty = "dashed")
lines(pre$year, pre$upperCI, type = "l", lty = "dashed")

lines(post$year, post$inci_modelled, type = "l")
lines(post$year, post$lowerCI, type = "l", lty = "dashed")
lines(post$year, post$upperCI, type = "l", lty = "dashed")

abline(v = c(2007), col = "red", lty = "dashed")
```

*Explain how to calculate effects (see Wagner et al. 2002)*













